A generalization of the Bradley-Terry model for draws in chess with an application to collusion

### 2020-10-11 inst/jebo_chess_revision2.tex
further tiny changes for JEBO

Bradley-Terry is not directly applicable to chess for two reasons:
firstly, the white player appears to be at a considerable advantage,
winning about 37.5\% of tournament games (black wins 27.6\%); and
secondly, many chess games, about 39.4\%, are drawn (figures from
Chessbase).  First-mover advantage is closely
  related to incumbent advantage, and the techniques presented here
  have been applied to ecological incumbency~\cite{hankin2010}.  More
  generally, \cite{szymanski2003} considers sporting contests as
  examples of competitive situations in which a contribution of some
  kind [effort, investment, attendance of the elite, etc] is elicited
  from contestants who may as a result win a prize.
  \citeauthor{szymanski2003} points out that this paradigm arguably
  covers executive corporate promotion, interviews for a position in
  an office, and most sporting tournaments.  As such, the phenomenon
  of collusion and indeed drawing in pairwise comparison is surely an
  interesting phenomenon.

@@

The seminal work of \citet{moul2009} presents
  sophisticated regression techniques to analyse draws and first-mover
  advantage.  They compare all-play-all events with knockout format
  events to test---and reject---hypotheses of a distinct Soviet
  style-of-play; the work I present also considers Soviet proclivity
  to draw, using data from a single event, specifically the Stockholm
  Interzonal tournament of (1962).

@@

The maximum likelihood estimate for the full 23-player dataset is
$\draw\simeq 0.043, \draw_s=0.354$; a constrained maximization gives a
likelihood ratio of about 5096, or a $p$-value of about $3.6\times
10^{-5}$: there is strong evidence that Soviet players do indeed draw
more frequently than non-Soviets.  Indeed we could go further and plot
a profile likelihood curve for the log-odds ratio
$\log\left(\draw_s/\draw\right)$ which, if Dirichlet, would be
accurately parabolic over a wide range~\cite[page 343]{ohagan2004}.
This is illustrated in figure~\ref{proflike}, that shows a credible
interval of about 1.8-2.4: we can be reasonably sure that
$\draw_s/\draw\geqslant e^{1.8}\simeq 6.05$.

\begin{figure}[htbp]
\begin{center}
\makebox{\includegraphics{collusion_chess_draws-proflike.pdf}}
\caption{Profile likelihood \label{proflike} for odds ratio
  $\log\left(\draw_s/\draw\right)$: note the accurately parabolic figure in
  the region plotted}
  \end{center}
\end{figure}

@@

\textcolor{DarkGreen}{The ``rest monster'' concept is due to a
  suggestion by an insightful JEBO referee.}

>>>>

Bradley-Terry is not directly applicable to chess for two reasons:
firstly, the white player appears to be at a considerable advantage,
winning about 37.5\% of tournament games (black wins 27.6\%); and
secondly, many chess games, about 39.4\%, are drawn (figures from
Chessbase).  First-mover advantage is closely
  related to incumbent advantage, and the techniques presented here
  have been applied to ecological incumbency~\citep{hankin2010}.  More
  generally, \cite{szymanski2003} considers sporting contests as
  examples of competitive situations in which a contribution of some
  kind [effort, investment, attendance of the elite, etc] is elicited
  from contestants who may as a result win a prize.
  \citeauthor{szymanski2003} points out that this paradigm arguably
  covers executive corporate promotion, interviews for a position in
  an office, and most sporting tournaments.  As such, the phenomenon
  of collusion and indeed drawing in pairwise comparison is surely an
  interesting phenomenon.

@@

The seminal work of \citet{moul2009} presents
  sophisticated regression techniques to analyse draws and first-mover
  advantage.  They compare all-play-all events with knockout format
  events to test---and reject---hypotheses of a distinct Soviet
  style-of-play; the work I present also considers Soviet proclivity
  to draw, using data from a single event, specifically the Stockholm
  Interzonal tournament of 1962.

@@

The maximum likelihood estimate for the full 23-player dataset is
$\draw\simeq 0.043, \draw_s=0.354$; a constrained maximization gives a
likelihood ratio of about 5096, or a $p$-value of about $3.6\times
10^{-5}$: there is strong evidence that Soviet players do indeed draw
more frequently than non-Soviets.  Indeed we could go further and plot
a profile likelihood curve for
$\log\left(\draw_s/\draw\right)$ which, if Dirichlet, would be
accurately parabolic over a wide range~\cite[page 343]{ohagan2004}.
This is illustrated in figure~\ref{proflike}, that shows a credible
interval of about 1.8-2.4: we can be reasonably sure that
$\draw_s/\draw\geqslant e^{1.8}\simeq 6.05$.

\begin{figure}[htbp]
\begin{center}
\makebox{\includegraphics{collusion_chess_draws-proflike.pdf}}
\caption{Profile likelihood \label{proflike} for 
  $\log\left(\draw_s/\draw\right)$: note the accurately parabolic figure in
  the region plotted}
  \end{center}
\end{figure}

@@

The ``rest monster'' concept is due to a suggestion by an insightful
JEBO referee.
